---
title: "Contamination Grader"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    df_print: kable
params:
  seed: 42
  bam: ""            # optional: bowtie2 BAM to grade (section 6)
  sample_type: "gut"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, fig.width = 7, fig.height = 3.5)
library(dplyr); library(tidyr); library(tibble); library(ggplot2)
set.seed(params$seed)
```

# 1. Inputs and assumptions

The grader takes one metagenomic sample's aligned reads and returns a **contamination score
from 0 (no evidence) to 1 (strong evidence)**, plus a per-taxon table explaining the score.

**Inputs:** for each aligned read, the reference genome (`ref_genome`), alignment position
(`pos`), mapping quality (`mapq`, MAPQ), alignment score (`aln_score`, bowtie2 `AS` tag), taxonomic
assignment (`taxon`, genus) and aligned length (`read_len`). Also each reference genome's length (`genomes`) and the `sample_type`
(`gut`, `skin` or `vaginal`). Each genus is assumed to have one reference genome; with
several, the coverage-evenness check would need to be run per reference.

**Two things this grader does not assume but are biologically relevant:**

- **No negative (blank) controls.** With blanks, frequency- and prevalence-based methods
  such as `decontam` are more powerful. See section 7.
- **No DNA concentration.** In low-biomass samples, contaminants make up a larger share of
  the reads. The abundance term below captures only part of that effect.

# 2. Method

Each taxon gets three contamination level values from 0 (looks genuine) to 1 (looks like contamination):

| Evidence | High when… |
|---|---|
| `f_habitat` | The genus is a reagent ("kitome") contaminant (0.95) or belongs to another body site (0.80). Native genera get 0.05, unknown genera 0.40. |
| `f_quality` | The worst of three read-level signals: reads align poorly (median score below 97% of the maximum); many reads map ambiguously (share with MAPQ < 20, i.e. they fit other references about as well); or reads pile up in a few spots instead of covering the genome evenly. |
| `f_abundance` | The taxon makes up very few reads (0.01% of reads scores 1; 10% or more scores 0). |

These combine into a per-taxon probability, and then a sample score (the chance that at
least one judged taxon is a contaminant):

$$\text{logit}(p_i) = -7 + 8 f_{habitat} + 3 f_{quality} + 1.5 f_{abundance}
\qquad \text{score} = 1 - \prod_{n_i \ge 10} (1 - p_i)$$

The weights are expert priors, not fitted values. Taxa with fewer than 10 reads are too thin to
judge.

**The niche database.** `habitat_table` is a two-column lookup: each row says a genus is
found in a habitat (`gut`, `skin`, `vaginal`), or is a known reagent contaminant (`kitome`).
A genus can appear under more than one habitat (e.g. *Prevotella*, *Streptococcus*). The
grader only asks one question of this table: for the sample's body site, is this genus
native, kitome, from another site, or not listed? That answer sets `f_habitat`.

```{r habitat}
habitat_table <- bind_rows(
  tibble(habitat = "gut", genus = c("Bacteroides", "Phocaeicola", "Faecalibacterium", "Roseburia",
         "Blautia", "Ruminococcus", "Bifidobacterium", "Akkermansia", "Alistipes",
         "Parabacteroides", "Escherichia", "Eubacterium", "Prevotella")),
  tibble(habitat = "skin", genus = c("Cutibacterium", "Staphylococcus", "Corynebacterium",
         "Micrococcus", "Streptococcus")),
  tibble(habitat = "vaginal", genus = c("Lactobacillus", "Gardnerella", "Fannyhessea", "Sneathia",
         "Megasphaera", "Mobiluncus", "Prevotella", "Streptococcus")),
  # reagent contaminants: Salter et al. 2014; Eisenhofer et al. 2019
  tibble(habitat = "kitome", genus = c("Ralstonia", "Bradyrhizobium", "Burkholderia",
         "Methylobacterium", "Sphingomonas", "Delftia", "Cupriavidus", "Stenotrophomonas",
         "Achromobacter", "Herbaspirillum", "Mesorhizobium", "Brevundimonas",
         "Pseudomonas", "Acinetobacter")))
```

# 3. The grader

The grader runs four assumption checks. A `FAIL` marks the score unreliable.

| Check | Assumption | Flags |
|---|---|---|
| C1 | The niche database covers this sample's taxa | Under 95% of reads (WARN) or under 80% (FAIL) come from genera in the table. Unknown genera would inflate the score. |
| C2 | The sample-type label is right | Another body site explains the reads better. A mislabelled sample looks contaminated. |
| C3 | Alignment scores are on the expected scale | Normalised scores fall outside 0–1 (e.g. bowtie2 end-to-end mode, where scores are ≤ 0). |
| C4 | Host reads were removed | Any reads labelled human (WARN, they are removed), or fewer than 1,000 microbial reads left (FAIL). |

**How `grade_contamination()` works, step by step** (the steps are marked in the code):

1. **Validate the input.** Stop with a clear message if a read column is missing or a
   reference has no genome length, since those would silently corrupt the score.
2. **Remove host reads.** `is_host()` recognises reads assigned to *Homo* or to a human
   reference (`GRCh38`, `hg38`, `chr1`…). They are dropped, and the count is kept for check C4.
3. **Summarise each taxon.** Group reads by genus and compute: number of reads, median
   alignment score as a fraction of the best possible score, share of reads with MAPQ < 20,
   and how many 1-kb genome windows the reads land in.
4. **Turn the summaries into suspicion values (0–1).**
   - `f_habitat` comes from the lookup table.
   - `f_quality` takes the worst of three read-level signals. The evenness signal compares
     the windows actually hit with the number expected if the same number of reads had
     landed at random: $W\,(1-(1-1/W)^{n})$ for a genome of $W$ windows and $n$ reads. A
     genuine genome gives a ratio near 1; reads piled into one spot give a ratio near 0.
   - `f_abundance` rises on a log scale as the taxon's share of reads falls from 10% to
     0.01%.
5. **Combine into a probability.** A logistic function turns the weighted sum into
   `p_contam`, the probability that the taxon is a contaminant.
6. **Run the assumption checks** (C1–C4) on the same data.
7. **Score the sample.** The noisy-OR gives the probability that at least one judged
   taxon is a contaminant. `contaminant_fraction` weights each taxon's reads by its
   `p_contam` to estimate *how much* of the sample is contamination.

The function returns a list: `score`, `contaminant_fraction`, `reliable` (no check failed),
`checks` and `taxa` (the per-taxon table, sorted by `p_contam`).

```{r grader}
clamp01 <- function(x) pmin(pmax(x, 0), 1)          # keep a value inside [0, 1]

# TRUE for reads from the human host (by taxon or by human reference name)
is_host <- function(r) r$taxon == "Homo" |
  grepl("^(chr([0-9]{1,2}|X|Y|MT?)$|GRCh|hg1[89]|hg38|CHM13|T2T)", r$ref_genome, ignore.case = TRUE)

grade_contamination <- function(reads, genomes, sample_type, table = habitat_table,
                                max_score_per_base = 2, window = 1000, min_reads = 10) {
  # Step 1: validate input
  need <- c("ref_genome", "pos", "mapq", "aln_score", "taxon", "read_len")
  if (length(miss <- setdiff(need, names(reads)))) stop("missing read columns: ", toString(miss))
  # Step 2: remove host reads (kept as `host` for check C4)
  host  <- is_host(reads)
  reads <- reads[!host, ]
  n     <- nrow(reads)
  if (length(nolen <- setdiff(reads$ref_genome, genomes$ref_genome)))
    stop(length(nolen), " reference(s) have no genome length, e.g. ", nolen[1])
  if (n == 0)   # nothing left to grade (C4 FAIL)
    return(list(score = NA_real_, contaminant_fraction = NA_real_, reliable = FALSE, taxa = NULL,
                checks = tibble(check = "C4 host reads removed", status = "FAIL",
                                detail = "all reads were human; no microbial reads to grade")))
  site_of <- function(site) table$genus[table$habitat == site]   # genera listed for a habitat

  # Step 3: one row per taxon with read-level summaries
  taxa <- reads |>
    left_join(genomes, by = "ref_genome") |>
    group_by(taxon) |>
    summarise(n_reads = n(), genome_len = first(genome_len),
              median_score = median(aln_score / (read_len * max_score_per_base)),
              low_mapq = mean(mapq < 20),
              windows_hit = n_distinct(pos %/% window), .groups = "drop") |>
    # Step 4: suspicion values, each 0 (genuine) to 1 (contaminant)
    mutate(
      f_habitat = case_when(taxon %in% site_of(sample_type) ~ 0.05,
                            taxon %in% site_of("kitome")    ~ 0.95,
                            taxon %in% table$genus          ~ 0.80,
                            TRUE                            ~ 0.40),
      # windows hit vs. expected if reads landed uniformly at random
      W          = pmax(1, ceiling(genome_len / window)),
      evenness   = clamp01(windows_hit / (W * (1 - (1 - 1 / W)^n_reads))),
      f_quality  = pmax(clamp01((0.97 - median_score) / 0.12),     # poor alignments
                        clamp01((low_mapq - 0.1) / 0.4),             # ambiguous mapping: 10% -> 0, 50% -> 1
                        clamp01(1 - evenness / 0.8)),                # patchy coverage
      f_abundance = clamp01(log10(0.1 / (n_reads / n)) / 3),      # 10% -> 0, 0.01% -> 1
      # Step 5: weighted sum -> probability via the logistic function
      p_contam   = plogis(-7 + 8 * f_habitat + 3 * f_quality + 1.5 * f_abundance)) |>
    arrange(desc(p_contam))

  # Step 6: assumption checks
  # share of reads from genera native to each site (C2), and from any listed genus (C1)
  share  <- sapply(c("gut", "skin", "vaginal"), \(s) sum(taxa$n_reads[taxa$taxon %in% site_of(s)]) / n)
  best   <- names(which.max(share))
  in_db  <- sum(taxa$n_reads[taxa$taxon %in% table$genus]) / n
  ns     <- reads$aln_score / (reads$read_len * max_score_per_base)   # normalised scores (C3)
  lvl    <- \(fail, warn) if (fail) "FAIL" else if (warn) "WARN" else "PASS"
  checks <- tribble(~check, ~status, ~detail,
    "C1 niche DB coverage", lvl(in_db < 0.8, in_db < 0.95),
      sprintf("%.1f%% of reads from genera in the DB", 100 * in_db),
    "C2 sample type fits",  lvl(best != sample_type & share[best] > 2 * share[sample_type],
                                best != sample_type | share[sample_type] < 0.5),
      sprintf("%.1f%% native to '%s'; best fit '%s'", 100 * share[sample_type], sample_type, best),
    "C3 score scale",       lvl(any(ns > 1.001) | median(ns) <= 0, median(ns) < 0.8),
      sprintf("median normalised score %.3f", median(ns)),
    "C4 host reads removed", lvl(n < 1000, any(host)),
      sprintf("%.1f%% human reads removed; %d microbial reads", 100 * mean(host), n))

  # Step 7: noisy-OR over taxa with enough reads to judge
  list(score = 1 - prod(1 - taxa$p_contam[taxa$n_reads >= min_reads]),
       contaminant_fraction = sum(taxa$p_contam * taxa$n_reads) / n,
       reliable = !any(checks$status == "FAIL"), checks = checks, taxa = taxa)
}
```

# 4. Validation on simulated samples

Each simulated sample has 15 native genera (even coverage, high alignment scores, mostly
MAPQ 60). It also has two spurious hits: poorly aligned, ambiguously mapped (MAPQ ≤ 10) reads
piled up in one spot, which are **not** contamination. Contaminated samples add three genera
from the kitome or from another body site. Their DNA is real, so it aligns well.

**How the simulator works.** `simulate_sample()` gives every genus in the lookup table a
random genome length (2–6 Mb). Its helper `draw()` produces `k` reads for one genus: evenly
spread positions with high scores and MAPQ 60 for genuine reads, or a tight cluster with
lower scores and low MAPQ for spurious hits. Native abundances are log-normal, and
`rmultinom()` splits the read budget among genera. Every read carries a `truth` label
(`native`, `spurious`, `contaminant`). The grader never sees it; it is only there to check
the results.

```{r simulate}
simulate_sample <- function(type, n = 10000, frac = 0, source = "kitome", len = 150) {
  genomes <- tibble(taxon = unique(habitat_table$genus), ref_genome = paste0(taxon, "_ref"),
                    genome_len = round(runif(length(taxon), 2e6, 6e6)))
  draw <- function(g, k, even = TRUE, q = 0.99, mq = 60) {
    L <- genomes$genome_len[genomes$taxon == g]
    tibble(ref_genome = paste0(g, "_ref"), taxon = g, read_len = len,
           pos = if (even) sample.int(L - len, k, TRUE) else round(runif(1, 1, L - 5000) + runif(k, 0, 3000)),
           aln_score = round(len * 2 * clamp01(rnorm(k, q, 0.01))),
           mapq = if (mq == 60) sample(c(60, 1), k, TRUE, c(.95, .05)) else sample(0:10, k, TRUE))
  }
  site    <- unique(habitat_table$genus[habitat_table$habitat == type])
  natives <- sample(site, min(15, length(site)))
  ab      <- rlnorm(length(natives), 0, 1.5)
  nat  <- bind_rows(Map(draw, natives, rmultinom(1, round(n * (1 - frac)), ab))) |> mutate(truth = "native")
  spur <- bind_rows(lapply(sample(natives, 2), draw, k = 30, even = FALSE, q = 0.86, mq = 0)) |> mutate(truth = "spurious")
  pool <- if (source == "kitome") habitat_table$genus[habitat_table$habitat == "kitome"] else
          setdiff(habitat_table$genus[!habitat_table$habitat %in% c(type, "kitome")], site)
  con  <- if (frac > 0) bind_rows(Map(draw, sample(unique(pool), 3), rmultinom(1, round(n * frac), c(.6, .3, .1)))) |>
            mutate(truth = "contaminant")
  list(reads = bind_rows(nat, spur, con), genomes = select(genomes, ref_genome, genome_len))
}
```

**Does the score separate clean from contaminated samples?** The code below simulates ten
samples for every combination of body site, contaminant source and contamination level,
grades each one, and reports the median and highest score per level:

```{r benchmark}
bench <- expand_grid(type = c("gut", "skin", "vaginal"), source = c("kitome", "other_site"),
                     frac = c(0, 0.001, 0.005, 0.02, 0.2), rep = 1:10) |>
  rowwise() |>
  mutate(score = { s <- simulate_sample(type, frac = frac, source = source)
                   grade_contamination(s$reads, s$genomes, type)$score }) |>
  ungroup()

bench |> group_by(source, `true contamination` = paste0(frac * 100, "%")) |>
  summarise(median_score = round(median(score), 3), max_score = round(max(score), 3), .groups = "drop") |>
  pivot_wider(names_from = source, values_from = c(median_score, max_score))
```

Clean samples score near 0 even with spurious hits, and 0.5% contamination is clearly
detected. At 0.1%, the contaminants usually have fewer than 10 reads each, so they aren't
judged. That is roughly the detection limit at 10,000 reads. Once a contaminant is found, the score is close to its
ceiling. It measures **strength of evidence, not amount**; for amount, use
`contaminant_fraction`.

**Does each assumption check catch its own failure?** The code below takes clean samples and
breaks one assumption at a time: it removes 6 of the 13 gut genera from the lookup table,
grades a vaginal sample as gut, replaces scores with bowtie2 end-to-end-style scores (≤ 0),
and adds human reads mapped to `GRCh38`. The table shows the score and which checks
fired:

```{r failures}
g   <- simulate_sample("gut")
v   <- simulate_sample("vaginal")
e2e <- g; e2e$reads$aln_score <- -round(rexp(nrow(g$reads), 1 / 6))   # bowtie2 end-to-end scale
hum <- g; nh <- round(nrow(g$reads) * 2 / 3)                            # 40% of reads human
hum$reads   <- bind_rows(g$reads, tibble(ref_genome = "GRCh38", taxon = "Homo", read_len = 150, mapq = 60,
                         pos = sample.int(3e9, nh), aln_score = 297))
hum$genomes <- bind_rows(g$genomes, tibble(ref_genome = "GRCh38", genome_len = 3.1e9))
gut_genera  <- habitat_table$genus[habitat_table$habitat == "gut"]
thin_db     <- filter(habitat_table, !(habitat == "gut" & genus %in% sample(gut_genera, 6)))

cases <- list(
  "None (baseline)"                   = grade_contamination(g$reads, g$genomes, "gut"),
  "Niche DB missing ~half of gut genera" = grade_contamination(g$reads, g$genomes, "gut", table = thin_db),
  "Vaginal sample labelled as gut"    = grade_contamination(v$reads, v$genomes, "gut"),
  "End-to-end alignment scores"       = grade_contamination(e2e$reads, e2e$genomes, "gut"),
  "40% human reads in the input"      = grade_contamination(hum$reads, hum$genomes, "gut"))

imap_dfr <- function(x, f) bind_rows(Map(f, x, names(x)))
imap_dfr(cases, \(r, nm) tibble(`broken assumption` = nm, score = round(r$score, 3),
  reliable = r$reliable,
  flagged = paste(with(r$checks, paste0(check, ": ", status)[status != "PASS"]), collapse = "; ")))
```

A missing niche database or a wrong label raises the score of a clean sample, and the
matching check flags it, so a high score with a C1 or C2 warning points to a data problem,
not contamination. Human reads are removed automatically, so they don't change the score,
but C4 reports them.

# 5. Worked example

A gut sample with 2% reagent contamination. The per-taxon table shows how each suspicion
value feeds into `p_contam`. The reagent genera rise to the top because of `f_habitat`, even
though their reads align well:

```{r example}
ex  <- simulate_sample("gut", frac = 0.02)
res <- grade_contamination(ex$reads, ex$genomes, "gut")
cat(sprintf("score = %.3f | estimated contaminant fraction = %.3f (true 0.020) | reliable = %s\n",
            res$score, res$contaminant_fraction, res$reliable))
res$taxa |> select(taxon, n_reads, low_mapq, f_habitat, f_quality, f_abundance, p_contam) |>
  mutate(across(where(is.double), \(x) signif(x, 2))) |> head(6)
```

# 6. Running on a real BAM

To grade real data, set the `bam` and `sample_type` parameters when knitting. The code
reads each primary aligned read's reference, position, length, MAPQ and `AS` tag with
Rsamtools, and takes genome lengths from the BAM header. The taxon is the genus at the start
of the reference name (`Bacteroides_fragilis_...`). If your taxonomy comes from elsewhere,
replace the `taxon` column. Human references (`chr1`, `GRCh38`…) are recognised and removed.
Tested on a simulated BAM: a gut sample with 2% reagent contamination and 300 human reads
scored 0.99, with the three reagent genera as the top suspects.

```{r bam, eval = nzchar(params$bam)}
x   <- Rsamtools::scanBam(params$bam, param = Rsamtools::ScanBamParam(
         what = c("rname", "pos", "qwidth", "mapq"), tag = "AS",
         flag = Rsamtools::scanBamFlag(isUnmappedQuery = FALSE, isSecondaryAlignment = FALSE)))[[1]]
hdr <- Rsamtools::scanBamHeader(params$bam)[[1]]$targets
reads <- tibble(ref_genome = as.character(x$rname), pos = x$pos, read_len = x$qwidth, mapq = x$mapq,
                aln_score = x$tag$AS, taxon = sub("_.*", "", ref_genome))
real <- grade_contamination(reads, tibble(ref_genome = names(hdr), genome_len = as.numeric(hdr)),
                            params$sample_type)
real$checks; real$score; head(real$taxa, 10)
```

# 7. Limitations

- **The weights are priors.** Fit them by logistic regression once labelled data exist
  (blanks, spike-ins, mock communities).
- **Blanks and DNA concentrations would help most.** Add the `decontam` prevalence method
  (taxon found more often in blanks than in samples) and frequency method (contaminant share
  rises as input DNA falls) as extra evidence terms.
- **Low-biomass samples** can be dominated by contaminants. The abundance term then points
  the wrong way, so reduce its weight.
- **The niche table is small** (5 skin and 8 vaginal genera). Replace it with a body-site
  catalogue such as HMP or GMrepo.
- **Only labelled human reads are removed.** Human reads assigned to a microbial reference
  are missed, so align to the human genome first.
